\textbf{On the Common Logical Structure of Classical and Quantum Mechanics}

### 2021-04-01 Articles/2020-QuantumLogic/Source/quantumlogic.tex
QLogic: review with Andrea

In this section we present our main argument, which is that the state space of quantum mechanics and the space of classical probability distributions possess a parallel structure that allows one to apply commutative propositional calculus (i.e. ``classical logic'') on the first and non-commutative propositional calculus (i.e. ``quantum logic'') on the second. That is, both theories will have two lattices of propositions: a commutative one that follows classical logic and a non-commutative one that follows quantum logic.

Note that the classical space we will be using is not phase-space itself, but the space of probability distributions over phase-space, which is a real Hilbert space and thus has a similar structure to the quantum counterpart. The central point is that many features of quantum logic stem not from quantum mechanics per-se, but from the fact that quantum mechanics is a probabilistic theory, which means a classical probabilistic theory will share the same features. Critically, we will see that the non-commutativity of the quantum logic lattice does not originate from non-commuting observables or superposition of quantum mechanics: it is a property of the lattice of subspaces of any vector space and, indeed, of the lattice of sugroups of any group more in general.

Mathematically, the classical commutative lattice will be provided by the Borel algebra, the smallest $\sigma$-algebra that contains the topology; the non-classical non-commutative lattice will be provided by the set of subgroups ordered by inclusion. We will see that the non-commutative lattice is a strict subset of the classical commutative one, meaning that everything that can be described in quantum logic can also be described in classical logic. Moreover, the commutative lattice will include propositions, like propositions on the expectation value of observables, that are of scientific interest.\footnote{In an experimental context, in fact, by ``measurement'' one typically refers to the statistical collection of many ``single takes''. Quantum logic describes the single-take measurements. For example, measurements of cross sections, the most typical type of measurement in the context of particle physics, are not part of the quantum lattice, and neither are measurements of the probabilities themselves.}

While topologies and $\sigma$-algebras are foundational structures in mathematics for differential geometry,[CITE: Lee, Introduction to Smooth Manifolds] measure theory and probability theory,[CITE: Grimmet/Stirzaker, Probability and Random Processes] they do not seem well-known among physicists and philosophers of science, therefore we briefly summarize their definition, use and meaning. [CITE: Kelly, Logic of Reliable Inquiry - Carcassi/Aidala, Assumptions of Physics]

We start with a set $X$ which can be thought to represent the set of all possible cases we need to distinguish: all quantum states in the quantum case; all possible probability distributions in the classical case; for a simpler example, the set of the real numbers for a continuous quantity. Both topologies and $\sigma$-algebras are collections of subsets of $X$. For example, the standard topology for the real numbers will contain all the open intervals $(a, b)$. The difference between a topology and a $\sigma$-algebra is what types of set operations one can take within them.

A collection of subsets may seem quite an abstract concept, but it can be readily made concrete: any statement on the objects represented by $X$ can be represented by a set. For example, the open set $(a, b)$ will exactly correspond to the statement ``the value of the quantity is between $a$ and $b$''. It is therefore useful to think about both the topology and the $\sigma$-algebra as collections of statements about $X$ instead of sets.

A $\sigma$-algebra is closed under countable union, countable intersection and complement with respect to $X$, meaning that performing those operations on the sets contained in a $\sigma$-algebra will return another set in the same $\sigma$-algebra. In terms of statements, this means that it is closed under countable conjunction, countable disjunction and complement: the $\sigma$-algebra can be thought as a Boolean algebra of statements. Note that this interpretation is exactly the one used in probability. In fact, a probability space is defined by three things: a sample space $\Omega$ which represents the set of all possible outcomes; a $\sigma$-algebra over $\Omega$ which represents all events, all propositions; and a measure that assigns a probability to each event.

A topology, instead, is closed under arbitrary union and finite intersection only. In terms of the statements, it is closed under arbitrary disjunction and finite conjunction. A topology is therefore not a Boolean algebra, but a Heyting algebra. Heyting algebras play a central role in intuitionist logic (and in computer science) as they describe lattices of provable (or computable) statements in finite time. In physics, we can understand the topology to represent statements that are experimentally verifiable in finite time.[CITE again Kelly and Carcassi/Aidala] For example, the open interval $(a, b)$ is an open set precisely because the statement ``the value of the quantity is between $a$ and $b$'' is something we can verify experimentally. Note that the sets in a topology are called open sets precisely because they are a generalization of the open intervals. On the other hand, a set with only one real value $\{ x_0 \}$ is not an open set, it is not in the topology, because the statement ``the value of the quantity is exactly $x_0$'' is not verifiable experimentally due to infinite precision.

Establishing a topology, then, means establishing what statements are assumed to be experimentally verifiable. Given a topology, its Borel algebra is the smallest $\sigma$-algebra that contains all sets in the topology. That is, it is the smallest Boolean algebra that contains all the verifiable statements associated to the topology. Topologies and $\sigma$-algebra are pervasive in physics, to the point that all physically interesting spaces are, directly or indirectly, topological spaces. And wherever we have a topology, we have a Borel algebra, which is a $\sigma$-algebra, a Boolean algebra.

On complete normed spaces, such as Hilbert spaces, there is a standard way to construct the topology. Let us see first how it works in a standard three-dimensional Euclidean space. Given two points $x$ and $y$, we can write the square of distance $\epsilon^2 = |y - x|^2$ as the square of the norm of the vector between the two points. The statement ``the position is within $\epsilon$ meters of $x$'' corresponds to the set $U = \{y \; | \; |y - x|^2 < \epsilon^2 \}$ of all the points whose distance from $x$ is less than $\epsilon$. These are exactly the open sets that generate the topology. This works in the same way for all normed spaces. A Hilbert space, in particular, comes with an inner product which also defines a norm. We can define a distance $\epsilon^2 = |\psi - \phi|^2=\langle \psi - \phi , \psi - \phi \rangle$ between two states, and the statement ``the state of the system is within $\epsilon$ of $\psi$'' corresponds to the set $U = \{\phi \; | \; |\psi - \phi|^2 < \epsilon^2\}$. As we'll see later, the space of classical probability distributions is also a Hilbert space where each point corresponds to the square root of the density. We will have a distance $\epsilon^2 = |\sqrt{\rho_1} - \sqrt{\rho_2}|^2=\langle \sqrt{\rho_1} - \sqrt{\rho_2} , \sqrt{\rho_1} - \sqrt{\rho_2} \rangle$ and we can proceed in the same exact way. In all normed spaces, the topology is constructed from sets of the form ``the object is within $\epsilon$ of $x$''.\footnote{In the case of Hilbert spaces and quantum mechanics, the map between topology and experimental verification does not work perfectly because absolute modulus and phase are not physically meaningful. This can be fixed by using the corresponding projective space, which removes the overdetermination. We are not going to do this as projective spaces have their own peculiarities and are used less.}

From this topology, one can construct the Borel algebra. The Borel algebra also contains the closed sets, which are defined to be the complement of open sets. Recall that, in quantum logic, the propositions corresponded to closed subspaces, which are closed sets and are therefore Borel sets. All the propositions of quantum logic are therefore readily found in the Borel algebra of the Hilbert space. The Borel algebra, which is a Boolean algebra that follows classical logic, contains also all the statements of quantum logic. This construction also applies identically for the classical cases, where we will also be able to talk about the set of all closed subspaces which will have a parallel structure.

While all of this may seem daunting to someone who has never seen this material, we want to stress that these are very standard constructions. We are not taking some obscure framework from some fringe field of mathematics: these are the standard mathematical techniques used to study these spaces and can be found in any textbook. What we are doing here is bringing out their full physical significance, which is too often overlooked. Understanding the link between the mathematical structures and the physics they represent is crucial to clearing out possible misunderstandings.

@@

For example, quantum mechanics can only be expressed as a probabilistic theory and therefore ALWAYS requires a structure that in classical mechanics is not mandatory. In quantum mechanics, all subspaces of dimension one represent a pure state while in classical mechanics this is not the case. If we take two intervals $U_x$ and $U_p$ in position and momentum, and form the statement ``the state is wholly in $U_x$ and $U_p$'', meaning that the vector components are zero outside of those regions, in the classical case we always find at least one $\rho(x,p)$ that satisfies the bounds, while in quantum mechanics, if the bounds violate the uncertainty principle, no suitable $\psi(x)$ can be found. The point is that the differences are in the content of the logical structure, on the content of the statements, not on the type of structure.

>>>>

In this section we present our main argument, which is that the state space of quantum mechanics and the space of classical probability distributions possess a similar structure that allows one to apply commutative propositional calculus (i.e. ``classical logic'') on the first and non-commutative propositional calculus (i.e. ``quantum logic'') on the second. That is, both theories will have two lattices of propositions: a commutative one that follows classical logic and a non-commutative one that follows quantum logic.

Note that the classical space we will be using is not phase-space itself, but the space of probability distributions over phase-space, which is a real Hilbert space and thus has a similar structure to the quantum counterpart. The central point is that many features of quantum logic stem not from quantum mechanics per-se, but from the fact that quantum mechanics is a probabilistic theory, which means a classical probabilistic theory will share the same features. We find that the main point of confusion stems from comparing quantum theory, which is inherently a probabilistic theory, to a non-probabilistic classical theory. This approach does not allow one to understand which differences are due to the classical/quantum nature of the theory, and which are due simply to the probabilistic/non-probabilistic nature. By comparing quantum theory to a probabilistic classical theory, we put ourselves in a position to fully understand which difference is due to what, and have a much clearer understanding of both.\footnote{Note that probabilistic theories are important even in classical mechanics. Due to the nature of experimental verification, even in classical mechanics, there are always going to be uncertainties in the preparation and measurements, and repeated measurements will always lead to statistical constructs.} Critically, we will see that the non-commutativity of the quantum logic lattice does not originate from non-commuting observables or superposition of quantum mechanics: it is a property of the lattice of subspaces of any vector space and, indeed, of the lattice of sugroups of any group more in general. [CITE: Introduction to Lattices and Order Theory, Davey, Priestley - 2002 CUP]

Mathematically, the commutative lattice will be provided in both classical and quantum mechanics by the Borel algebra, the smallest $\sigma$-algebra that contains the topology; the non-commutative lattice will be provided by the set of subgroups ordered by inclusion. We will see that the non-commutative lattice is a strict subset of the commutative one, meaning that everything that can be described in quantum logic can also be described in classical logic. Moreover, the commutative lattice will include propositions, like propositions on the expectation value of observables, that are of scientific interest.\footnote{In an experimental context, in fact, by ``measurement'' one typically refers to the statistical collection of many ``single takes''. Quantum logic describes the single-take measurements. For example, measurements of cross sections, the most typical type of measurement in the context of particle physics, are not part of the quantum lattice, and neither are measurements of the probabilities themselves.}

%they do not seem well-known among physicists and philosophers of science, therefore we briefly summarize their definition, use and meaning. [CITE: Kelly, Logic of Reliable Inquiry - Carcassi/Aidala, Assumptions of Physics]

%It is useful, at this point, to review $\sigma$-algebras, which are foundational structures in mathematics,

Note that mathematical branches important in the context of physics, like measure theory [CITE: Measure theory, by Donald L. Cohn] and probability theory,[CITE: Grimmet/Stirzaker, Probability and Random Processes] are built not top of $\sigma$-algebras, which should be considered foundational structure. Let us therefore review some crucial aspects.

We start with a set $X$. In the quantum case, this will correspond to the set of all possible quantum states. In the classical case, this will corrsepond to the set of all possible probability distributions over phase space. We should stress, however, that $\sigma$-algebras are defined independently of what the elements of $X$ actually are. For a simpler example, $X$  could correspond to the set of the real numbers for a continuous quantity. A $\sigma$-algebras is a collection of subsets of $X$, closed under countable union, countable interesection and complement with respect to $X$, meaning that performing those operations on the sets contained in a $\sigma$-algebra will return another set in the same $\sigma$-algebra. For example, a standard $\sigma$-algebras for the real numbers will typically contain all the open intervals $(a, b)$.

A collection of subsets may seem quite an abstract concept, but it can be readily made concrete: any statement on the objects represented by $X$ can be represented by a set. For example, the open set $(a, b)$ will exactly correspond to the statement ``the value of the quantity is between $a$ and $b$''. It is therefore useful to think about $\sigma$-algebras as collections of statements about $X$ instead of sets. As the $\sigma$-algebra is closed under countable union, countable intersection and complement with respect to $X$, the corresponding statements will be closed under countable conjunction, countable disjunction and complement: the $\sigma$-algebra can be thought as a Boolean algebra of statements. Note that this interpretation is exactly the one used in probability. In fact, a probability space is defined by three things: a sample space $\Omega$ which represents the set of all possible outcomes; a $\sigma$-algebra over $\Omega$ which represents all events, all propositions; and a measure that assigns a probability to each event.

Every topological space comes already equipped with a $\sigma$-algebra called Borel algebra. A topology, in fact, is also a collections of sets, called open sets, and the Borel algebra is defined to be the smallest $\sigma$-algebra that contains all the open sets. The topology will not be directly important for us, but only as a mean to construct the Borel algebra. Complete normed spaces, such as Hilbert spaces, have a canonical way to construct a topology and therefore a $\sigma$-algebra. Let us see first how it works in a standard three-dimensional Euclidean space. Given two points $x$ and $y$, we can write the square of distance $\epsilon^2 = |y - x|^2$ as the square of the norm of the vector between the two points. The statement ``the position is within $\epsilon$ meters of $x$'' corresponds to the set $U = \{y \; | \; |y - x|^2 < \epsilon^2 \}$ of all the points whose distance from $x$ is less than $\epsilon$. These are exactly the open sets that generate the topology, and the Borel algebra will be the smallest $\sigma$-algebra that contains these sets. This works in the same way for all normed spaces. A Hilbert space, in particular, comes with an inner product which also defines a norm. We can define a distance $\epsilon^2 = |\psi - \phi|^2=\langle \psi - \phi , \psi - \phi \rangle$ between two states, and the statement ``the state of the system is within $\epsilon$ of $\psi$'' corresponds to the set $U = \{\phi \; | \; |\psi - \phi|^2 < \epsilon^2\}$. As we'll see later, the space of classical probability distributions is also a Hilbert space where each point corresponds to the square root of the density.\footnote{The choice of the square root of the density over the density itself is really a technical choice. Of all the $L^p$ spaces, the only Hilbert space is $L^2$, the space of the square-integrable functions.} We will have a distance $\epsilon^2 = |\sqrt{\rho_1} - \sqrt{\rho_2}|^2=\langle \sqrt{\rho_1} - \sqrt{\rho_2} , \sqrt{\rho_1} - \sqrt{\rho_2} \rangle$ and we can proceed in the same exact way. In all normed spaces, the topology is constructed from sets of the form ``the object is within $\epsilon$ of $x$''.\footnote{In the case of Hilbert spaces and quantum mechanics, the map between topology and experimental verification does not work perfectly because absolute modulus and phase are not physically meaningful. This can be fixed by using the corresponding projective space, which removes the overdetermination. We are not going to do this as projective spaces have their own peculiarities and are used less.}

It is a simple result that the Borel algebra thus constructed contains all propositions of quantum logic.

\begin{prop}
	Let $H$ be a Hilbert space. Let $\Sigma$ be the Borel $\sigma$-algebra induced by its inner product. Let $L(H)$ be the lattice of closed subspaces of $H$. Then $L(H) \subset \Sigma$.
\end{prop}

Proof. Let $p \in L(H)$ be a closed sub-space of $H$. Then $p$ is a closed set, the complement of an open set, and therefore a Borel sets: $p \in \Sigma$. Let $v \in H$ be a vector in the Hilbert space and consider the singleton $U = \{ v \}$. This is not a closed sub-space and therefore $U \notin L(H)$. As the topology of a Hilbert space is Hausdorff, all singletons are closed and therefore $U \in \Sigma$. This means $L(H) \subset \Sigma$. This concludes the proof.

In quantum logic, the propositions corresponded to closed subspaces, which are closed sets and are therefore Borel sets. All the propositions of quantum logic are therefore readily found in the Borel algebra of the Hilbert space. The Borel algebra, which is a Boolean algebra that follows classical logic, contains also all the statements of quantum logic. This construction also applies identically for the classical cases, where we will also be able to talk about the set of all closed subspaces which will have a parallel structure.

While all of this may seem daunting to someone who has never seen this material, we want to stress that these are very standard constructions. We are not taking some obscure framework from some fringe field of mathematics: these are the standard mathematical techniques used to study these spaces and can be found in any textbook. [CITE: Elements of Hilbert Spaces and Operator Theory, Harkrishan Lal Vasudeva, https://www.springer.com/gp/book/9789811030192 ] What we are doing here is bringing out their full physical significance, which is too often overlooked. Understanding the link between the mathematical structures and the physics they represent is crucial to clearing out possible misunderstandings.

@@

For example, quantum mechanics can only be expressed as a probabilistic theory and therefore ALWAYS requires a structure that in classical mechanics is not mandatory. In quantum mechanics, all subspaces of dimension one represent a pure state while in classical mechanics this is not the case. If we take two intervals $U_x$ and $U_p$ in position and momentum, and form the statement ``the state is wholly in $U_x$ and $U_p$'', meaning that the vector components are zero outside of those regions, in the classical case we always find at least one $\rho(x,p)$ that satisfies the bounds, while in quantum mechanics, if the bounds violate the uncertainty principle, no suitable $\psi(x)$ can be found. The point is that the differences are in the content of the logical structure, on the particular relationships among the particular statements, not on the type of structure.
